Verification of a TCP/IP stack

### 2020-09-07 paper/paper.tex
rework intro

    In the development of the Internet, there is a protocol that has a very special place: TCP.
    It has been described by Vincent Cerf and Bob Kahn who can be considered as the fathers for the Internet
    for their work, and the protocol has later been standardized in a RFC norm~\cite{rfc793}.
    Since this date, the TCP has been widely used on the Internet as the underlying protocol for application layers such as HTTP,
    FTP or TLS. TCP uses itself an underlying protocol to work, the Internet layer IP protocol which is itself
    based on a link layer, such as Ethernet. This stacking of protocols is referred to as the TCP/IP stack.

    The last decade has seen a growing interest for security and privacy in all the fields where computers are
    involved. The development of safe applications thanks to static analyzers and formal methods, that was first
    restricted to critical embedded software with the Astr\'ee analyzer~\cite{cousot2005astree} or the B method, are
    now adopted on a larger scale. New tools have been developed to promote the use of formal methods. Among them we
    can mention the SMT-solvers Z3~\cite{de2008z3} and Alt-Ergo~\cite{conchon:hal-01960203},
    the higher level language F*~\cite{mumon} that can prove assertions
    using these SMT-solvers or the more automatic static analyzer by abstract interpretation Frama-C~\cite{cuoq2012frama}
    that can help to identify undefined behaviors in C programs. Theorem provers, such as Coq or Isabelle, are another class
    of tools, where proofs are done manually.

    In the domain of networking and cryptography, these tools have been used to make secure implementations.
    Recently, Reiher and al have proposed a domain-specific language, RecordFlux~\cite{Reiher2019RecordFluxFM},
    to specify the message formats for real world protocols, that is written in SPARK.
    This is also the goal followed by the Everest project~\cite{bhargavan2017everest}.
    The authors implemented cryptographic functions such as AES or
    SHA2 as well as TLS, the cryptographic protocol that provides security
    to HTTPS and is its heart. The implementation is written
    in F* and is a combination of several sub-project such as miTLS~\cite{bhargavan2013implementing}
    or HACL*~\cite{zinzindohoue2017hacl}. The weakness of the Everest project lies in its need to use the untrusted
    transport layer TCP. If a bug exists in the transport layer, it can also jeopardize the higher level layers.

    The importance of the TCP protocol in the network communications has driven some people to work
    on it and try to ensure the safety of the protocol. Among these works, in~\cite{smith1996formal}
    the author has tried to formally verify that a TCP implementation verify some specifications.
    Another work comes from Tom Ridge and al~\cite{ridge2008rigorous} who have implemented the TCP
    protocol in the HOL proof assistant.
    But only few works has been done. Doing formal verification on a TCP implementation face up to two
    problems. The first one is the complexity of the TCP protocol that is written to be concurrent.
    The second problem comes from the norm itself that is written in a natural, informal language: English.
    Then a lot of part are under specified and extracted strong specifications can be hard.
    As a result, it does not exists a safe implementation for TCP that can be used in the industry.

    Our approach has been to consider an existing implementation of a TCP/IP stack and replace and integrate
    some bricks by Ada/SPARK code.
    We have chosen the Oryx Embedded's library, CycloneTCP\footnote{\url{https://www.oryx-embedded.com/products/CycloneTCP.html}}.
    The library is written in C, which is suitable to interface with Ada/SPARK, and is developed to run
    on embedded processors. We focused our work on rewriting in SPARK the TCP protocol because of the importance
    it has in the network communication as well as the socket API to provide a safe interface to the programmers
    to use our safe TCP implementation.

    In the first section we explore more in details the TCP protocol and the state machine that governs the behavior of the connection.
    Then in the next section we explain how we have formalized and implemented in SPARK the TCP user functions and the socket API.
    Finally in the last section we present some results of our work, like the bugs that have been found and eliminated thanks to our
    new implementation.

@@

\section{Specification Techniques Used}

@@

\section{Verification of the TCP protocol and the Socket API}

@@

\section{Results and bugs found}

>>>>

    In the development of the Internet, the TCP protocol that has a very special place.
    Initially described by the ``fathers of the Internet'' Vincent Cerf and Bob Kahn, the TCP protocol was standardized in 1981 by RFC 793~\cite{rfc793}.
    Since this date, TCP has been widely used on the Internet as the underlying protocol for application layers such as HTTP,
    FTP or TLS. TCP is based on the underlying IP protocol which is itself
    based on a link layer such as Ethernet. This stacking of protocols is referred to as the TCP/IP stack.

    Given the importance of the TCP protocol in network communications, some researchers have attempted to formalize the protocol to either prove key properties~\cite{smith1996formal}
    or even to fully prove the correctness of an implementation against a complete specification~\cite{ridge2008rigorous}.
    This latter implementation was performed in the HOL proof assistant, and while the authors note that it would be possible in theory to extract an implementation in Haskell from
    this work, this has not been adopted in practice.
    The formalization of TCP specification and verification of a TCP implementation face two difficulties:
    RFC 793 is written in English, leaving many parts underspecified; the TCP protocol is inherently concurrent which makes it complex to specify and verify.
    As a result, there is currently no formally verified implementation for TCP that is usable in industry.

    Other communication protocols have been formalized in recent years, leading to a formally verified implementation.
    The most prominent project in this area is the Everest project~\cite{bhargavan2017everest}, which led to a formally verified implementation in F* of TLS called miTLS~\cite{bhargavan2013implementing}.
    Note that miTLS still relies on an untrusted implementation of TCP, and thus could benefit from the results of our work.
    The approach taken in project Everest was to reimplement the formalized protocols completely in F*, a research programming language targeted at formal program verification.
    This approach is often difficult to adopt in industry, as codebases are evolved incrementally rather than replaced, both for economical and practical reasons.

    Our approach in this work has been to incrementally evolve an existing TCP/IP stack in C by replacing parts of the code with formally verified code in SPARK.
    The stack chosen, CycloneTCP\footnote{\url{https://www.oryx-embedded.com/products/CycloneTCP.html}}, is targeted at embedded processors with no Operating System.
    In Section~\ref{sec:TCP} we describe the TCP protocol, in particular the state machine that governs the behavior of a connection.
    In Section~\ref{sec:spec} we present specificities of the specification technique that we used, and in Section~\ref{sec:verif} we explain how we formalized and implemented in SPARK the TCP user functions and the underlying socket API.
    In Section~\ref{sec:results} we present the results of our work, in particular bugs that have been found and eliminated thanks to our approach, before we conclude.

@@

\section{Specification Techniques Used}
\label{sec:spec}

@@

\section{Verification of the TCP protocol and the Socket API}
\label{sec:verif}

@@

\section{Results and bugs found}
\label{sec:results}
